Relative parabolic factorizations of Chevalley groups

### 2015-07-31 gauss.tex
Rewritten paragraph about Gauss decompositions

\begin{thm}
Assume that for every terminal subsystem $\Delta$ the corresponding relative elementary subgroup admits the following decomposition:
\[ \E(\Delta,R,I)=\Hh(\Delta,R,I)\U^+(\Delta,I)\ldots\U^\pm(\Delta,I) \]
with $N$ triangular factors. Then the ambient relative elementary subgroup admits a decomposition
\[ \E(\Phi,R,I)=\Hh(\Phi,R,I)\U^+(\Phi,I)\ldots\U^\pm(\Phi,I) \]
of the same length $N$.
\end{thm}
\begin{proof}
Consider a set $Y=\Hh(\Phi,R,I)\U^+(\Phi,I)\ldots\U^\pm(\Phi,I)$. To show that $Y=\E(\Phi,R,I)$ one has to show that
\begin{enumerate}
\item $XY\subseteq Y$ for some set $X$ that generates $E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$;
\item $Y^{\E(\Phi,R)}\subseteq Y$, so that $Y$ is indeed normalized by $E(\Phi,R)$.
\end{enumerate}
We will take $X=\left\{x_\alpha(\xi)\mid\alpha\in\Pi,\ \xi\in I \right\}$. The set $X^{\E(\Phi,R)}$ contains elements $x_\alpha(\xi)$ for all $\alpha\in\Phi$ (since every root is conjugated to a fundamental one under the action of Weyl group, and the torus normalizer is contained in $\E(\Phi,R)$). Therefore $\E(\Phi,I)\subseteq\langle X\rangle^{\E(\Phi,R)}$ Now obviously $XY\subseteq Y$, since $\Hh(\Phi,R,I)$ normalizes each of $\X_\alpha(I)$, and for any $\alpha\in\Phi^+$ one has $\X_\alpha(I)\in\U(\Phi,I)$.

As for the second part, we use the fact that $\E(\Phi,R)$ is generated by $\left\{x_\alpha(\xi)\mid\alpha\in\pm\Pi,\ \xi\in R\right\}$. Fix a root $\alpha\in\pm\Pi$. It falls into one of the terminal subsystems, say $\Delta$. We expand $Y$ as follows:
\[ Y=\Hh(\Phi,R,I)\cdot\U(\Delta,I)\ldots\U^\pm(\Delta,I)\cdot\U(\Sigma,I)\ldots\U^\pm(\Sigma,I). \]
Here $\Sigma=\Phi^+\setminus\Delta$. Now
\begin{multline*}
x_\alpha(\xi)\cdot Y\cdot x_\alpha(-\xi) = x_\alpha(\xi)\cdot\E(\Delta,R,I)\cdot\U(\Sigma,I)\ldots\U^\pm(\Sigma,I)\cdot x_\alpha(-\xi) =\\
= x_\alpha(\xi)\cdot\E(\Delta,R,I)\cdot x_\alpha(-\xi)\cdot\U(\Sigma,I)\ldots\U^\pm(\Sigma,I) = Y.\qedhere
\end{multline*}
\end{proof}

>>>>

\begin{thm}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1, \Delta_\ell$ be
two reductive subsystems of $\Phi$ corresponding to endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1,\ell$ admit unitriangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I), i=1,\ell. \]
Then the group $\E(\Phi, R, I)$ admits decomposition with the same number of factors:
\begin{equation} \nonumber\E(\Phi, R, I) = \Hh(\Phi,R,I) \cdot \U(\Phi^+,I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm,I). \end{equation}
\end{thm}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi,R,I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi,R)$, i.\,e. $Y^{\E(\Phi,R)}\subseteq Y$;
\item there exists a set $X$ generating $\E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for $\alpha\in \pm \Pi$, $\xi\in R$.
Take any root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or $i=\ell$ and we can expand $Y$ as
\[ Y=\Hh(\Phi,R,I) \cdot \U(\Delta_i^+,I) \cdot \ldots \cdot \U^\pm(\Delta_i,I) \cdot \U(\Sigma_i^+,I) \cdot \ldots \cdot \U(\Sigma_i^\pm,I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has $h^{x_\alpha(\xi)} = h \cdot x_\alpha(\xi)^h x_\alpha(-\xi) = h \cdot x_\alpha(s)$ for some $s\in I$, therefore by the assumption of the theorem
\begin{multline*}
Y^{x_\alpha(\xi)} \subseteq \Hh(\Phi,R,I) \cdot X_{\alpha}(I) \cdot \E(\Delta_i,R,I)^{x_\alpha(\xi)} \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) =\\
= \Hh(\Phi, R, I) \cdot \E(\Delta_i,R,I) \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) = Y.
\end{multline*}

Now set $X=\left\{x_\alpha(\xi)\mid\alpha\in\Pi,\ \xi\in I \right\}$. 
Every root is a conjugate of a simple root under the action of $W(\Phi)$.
Since $\E(\Phi,R)$ contains the normalizer of the torus, the set $X^{\E(\Phi,R)}$ contains all the generators of the group $\E(\Phi, I) = \langle x_\alpha(\xi)$, $\alpha\in\Phi, \xi\in I \rangle$.
Finally, inclusion $XY \subseteq Y$ follows from the fact that $\Hh(\Phi,R,I)$ normalizes every root subgroup $X_\alpha(I)$.

\end{proof}
